Left-linear rules in adhesive categories

### 2024-02-12 icalp/fusione.tex
examples
minima

\documentclass[a4paper,UKenglish,cleveref,pdftex, thm-restate,numberwithinsect,anonymous]{lipics}
\newif\ifreport
\reporttrue
%\reportfalse

@@

There is, however, a further problem.  For linear
rewriting systems, switch equivalence can be characterised in terms of the
existence of what is called a consistent permutation~\cite{xxx}.
%(roughly, an isomorphism between their colimits). 
This result, which
is folklore for various categories of graphs, is explicitly proved
here for $\mathcal{M}$-adhesive categories.
%
The characterisation of switch equivalence based on consistent
permutations is the key to derive some
properties for linear rewriting systems, namely
\begin{enumerate}
\item \emph{confluence of switching}: when transforming
  sequences of rewriting steps into equivalent ones by switching independent steps, the
  result not depends on the order in which switchings are applied, but
  only on the resulting permutation;

\item \emph{globality of independence}: if two independent steps are
  moved forward or backward due to the application of switchings to a
  derivation, they remain independent.
\end{enumerate}
Properties (1) and (2) allow one to deduce a canonical form for equivalence proofs between derivations seen as concurrent computations.

Unfortunately, for left-linear rewriting systems the characterisation
of switch equivalence based on consistent permutations fails: the
existence of a consistent permutation is only necessary but not
sufficient, hence it cannot be used for proving (1) and (2).
%
This leads to the question asking whether the properties above hold.
We show that actually much of the theory can be retained: confluence
of switching still holds, while globality of independence holds in a
weak form, conceptually linked to the fact that fusions introduce a
form of disjunctive causality. Finally, a canonical form for switching
sequences can also be recovered.

@@

This first section is devoted to recall the definition and the basic theory of \emph{$\mathcal{M}$-adhesive categories} \cite{azzi2019essence,ehrig2012,ehrig2014adhesive,lack2005adhesive,heindel2009category}.

@@

\begin{example}
  \label{ex:adhesive}
  Every topos is adhesive~\cite{lack2006toposes}, thus in particular
  $\cat{Set}$ is adhesive. By the closure properties recalled before,
  every presheaf $[\cat{X},\cat{Set}]$ is adhesive and in particular
  the category $\cat{Graph} = [ E \rightrightarrows V, \cat{Set}]$ is
  adhesive where $E \rightrightarrows {V}$ is the two object category
  with two morphisms $s,t \colon{E} \to {V}$. In a similar way,
  various categories of hypergraphs can be shown to be adhesive, 
  such as term graphs and hierarchical graphs~\cite{CastelnovoGM24}.
% while the category $\cat{sGraphs}$ of simple graphs, i.e., graphs without
%  parallel edges is $\reg{(\cat{sGraphs})}$-adhesive.\todo{Add reference for simple graphs, add
 %   hierarchical graphs and term
 %   graphs\cite{castelnovo2023thesis,CastelnovoGM22}).}
\end{example}

@@

In this section we discuss the notion of independence between
consecutive rewriting steps, a fundamental ingredient of the rewriting
theory, which comes into play when viewing a derivation sequence as a
concurrent computation. We observe that moving from linear to
left-linear rules leads to the failure of various basic properties of
the classical notion of independence used in the linear setting and we
single out a framework in which these can be re-established, possibly
in a wekened form.

@@

\subsection{well switching rewriting systems}\label{subsec:verytame}

@@

\begin{example}
  Consider the rewriting system used as running example (rules in
  Fig.~\ref{fi:rules}. Assume to add a fourth rule $\rho_4$ as in
  Fig.~\ref{fi:rule-cas} and consider the derivation.
\end{example}

>>>>

\documentclass[a4paper,UKenglish,cleveref,pdftex,thm-restate,numberwithinsect,anonymous]{lipics}
\newif\ifreport
\reporttrue
%\reportfalse

@@

There is, however, a further problem.  For linear
rewriting systems, switch equivalence can be characterised in terms of the
existence of what is called a consistent permutation~\cite{xxx}.
%(roughly, an isomorphism between their colimits). 
This result, which
is folklore for various categories of graphs, is explicitly proved
here for $\mathcal{M}$-adhesive categories.
%
The characterisation of switch equivalence based on consistent
permutations is the key to derive some
properties for linear rewriting systems, namely
\begin{enumerate}
\item \emph{consistency of switching}: when transforming
  sequences of rewriting steps into equivalent ones by switching independent steps, the
  result not depends on the order in which switchings are applied, but
  only on the resulting permutation;
\item \emph{globality of independence}: if two independent steps are
  moved forward or backward due to the application of switchings to a
  derivation, they remain independent.
\end{enumerate}
Properties (1) and (2) allow one to deduce a canonical form for equivalence proofs 
between sequences of rewriting steps seen as concurrent computations.

Unfortunately, for left-linear rewriting systems the characterisation
of switch equivalence based on consistent permutations fails: the
existence of a consistent permutation is only necessary but not
sufficient, hence it cannot be used for proving (1) and (2).
%
This leads to the question asking whether the properties above hold.
We show that actually much of the theory can be retained: consistency
of switching still holds, while globality of independence holds in a
weak form, conceptually linked to the fact that fusions introduce a
form of disjunctive causality. Finally, a canonical form for switching
sequences can also be recovered.

@@

This first section recalls definitions and basic theory of \emph{$\mathcal{M}$-adhesive categories} \cite{azzi2019essence,ehrig2012,ehrig2014adhesive,lack2005adhesive,heindel2009category}.

@@

\begin{example}
  \label{ex:adhesive}
  Every topos is adhesive~\cite{lack2006toposes}, thus in particular
  $\cat{Set}$ is adhesive. By the closure properties recalled before,
  every presheaf $[\cat{X},\cat{Set}]$ is adhesive and in particular
  the category $\cat{Graph} = [ E \rightrightarrows V, \cat{Set}]$ is
  adhesive where $E \rightrightarrows {V}$ is the two object category
  with two morphisms $s,t \colon{E} \to {V}$. In a similar way,
  various categories of hypergraphs can be shown to be adhesive, 
  such as term graphs and hierarchical graphs~\cite{CastelnovoGM24},
  as well as the category $\cat{sGraphs}$ of simple graphs, i.e. graphs without
  parallel edges, is $\reg{(\cat{sGraphs})}$-adhesive~\cite{xxx}
.%\todo{Add reference for simple graphs, add
 %   hierarchical graphs and term
 %   graphs\cite{castelnovo2023thesis,CastelnovoGM22}).}
\end{example}

@@

In this section we discuss the notion of independence between
consecutive rewriting steps, a fundamental ingredient of the rewriting
theory, which comes into play when viewing a sequence of steps as a
concurrent computation. We observe that moving from linear to
left-linear rules leads to the failure of various basic properties of
the classical notion of independence used in the linear setting and we
single out a framework in which these can be re-established, possibly
in a wekened form.

@@

\subsection{Well-switching rewriting systems}\label{subsec:verytame}

@@

\begin{example}
  Consider the rewriting system used as running example (rules in
  Fig.~\ref{fi:rules}. Assume to add rules $\rho_3$ and $\rho_4$ as
  depicted in Fig.~\ref{fi:rule-cas}. Consider the derivation in
  Fig.~\ref{fi:derF}
  \begin{figure}
    \input{Figure/rules-cas.tex}    
    \caption{Additional rules.}
    \label{fi:rules-cas}
  \end{figure}

  \begin{figure}
    \input{Figure/derF.tex}    
    \caption{A derivation $\der{F}$.}
    \label{fi:derF}
  \end{figure}
\end{example}
